WASP-4 is Accelerating Towards the Earth

### 2020-01-31 paper/ms.tex
minor updates

A cursory analysis of the new HIRES observations led to our detection
of a linear trend in the residuals after fitting out the orbit of
WASP-4b.  This prompted us to acquire speckle images using Zorro at
Gemini-South \citep[see][and the instrument
web-pages\footnote{\url{www.gemini.edu/sciops/instruments/alopeke-zorro/}}]{scott_nessi_2018}.
Zorro is a dual-channel speckle interferometer with narrow-band
filters centered at 562$\,$nm and 832$\,$nm.

% TODO: might Howell suggest a paragraph for how the contrast ratios
% were derived?
We observed WASP-4 twice, on the night of September 11-12 and also
on the night of September 28-29. 
{\bf We reconstructed the images following...}
{\bf We reduced the reconstructed speckle images to contrast curves by
injecting and recovering point-sources...}
The contrast curves showed that the second night, which had better
seeing, also produced the more constraining result.  The 832$\,$nm
results were also the most useful, since faint companions are redder than
the host star.  We therefore opted for the 832$\,$nm September 28-29
results for the remaining analysis.

@@

The system's acceleration towards our line-of-sight causes a decrease
in the apparent orbital period.  The period derivative expected from
radial velocities is
\begin{align}
  \dot{P}_{\rm\,RV}
  = \frac{\dot{v}_{\rm r} P}{c}
  = -5.94 \pm 0.39~{\rm msec}\,{\rm yr}^{-1}.
\end{align}
In other words, the majority of the period decrease observed in
transits ($\dot{P}= - 8.64 \pm 0.89~{\rm msec}\,{\rm yr}^{-1}$) seems to
be explained by the acceleration of the host star.

@@

Given a linear radial velocity trend, we can place lower-limits on the
mass and semi-major axis of additional bodies in the system.  For a
quick estimate of the minimum mass required to explain the linear
trend in WASP-4, we turn to \citet{feng_california_2015}.  As they
discuss, the scenario that yields the minimum companion mass for a
system with a linear trend is a companion with $e\approx0.5$ and
$\omega=90^\circ$.  Substituting $P\approx 1.25\tau$, $K \approx
0.5\tau \gammadot$ into the mass function \citep[{\it
e.g.},][]{wright_efficient_2009} yields
\begin{equation}
 M_{\rm min} \approx 5.99\,M_{\rm Jup}
  \left( \frac{\tau}{{\rm yr}} \right)^{4/3}
  \left| \frac{\dot{\gamma}}{{\rm m\,s^{-1}\,day^{-1}}} \right|
  \left( \frac{M_\star}{{M_\odot}} \right)^{2/3},
\end{equation}
where $\tau$ is the observing baseline.  For WASP-4, this gives
$M_{\rm min} = 4.9 M_{\rm Jup}$.  Higher masses are allowed for
companions that orbit further from the star: at fixed $\dot{\gamma}$,
$M_{\rm comp} \propto a^2$
\citep{torres_substellar_1999,liu_crossing_2002}.

@@

Due to the custom filters of the Zorro imager, and corresponding lack
of synthetic photometry, we further assumed that all sources had
blackbody spectra. While this is a bold simplification, we do not
readily have access to the planetary and stellar atmosphere models
needed for the consistent calculation with the \texttt{COND03} and
\texttt{MESA} models.  We therefore adopted the effective temperatures
and bolometric luminosities from the \citet{baraffe_evolutionary_2003}
and MIST isochrones.  Using these theoretical quantities and the
empirically-measured Zorro bandpasses, we calculated absolute
magnitudes in the 562 and 832 nm Zorro bands for stellar and planetary
mass companions.  Applying the same calculation to WASP-4 itself using
the effective temperature and bolometric luminosity from
\citetalias{bouma_wasp4b_2019}, we derived the transformation from
contrast ratio to companion mass.  The resulting limits are shown in
Figure~\ref{fig:mass_sma}.

@@

We assigned each companion a mass and semimajor axis from log-uniform
distributions within the grid cell. We drew the inclination from a
uniform distribution in $\cos i$.  For companion masses less than
$10\,M_{\rm Jup}$, we drew the eccentricity from
\citet{kipping_beta_2013}'s long-period exoplanet Beta distribution
($a=1.12$, $b=3.09$).  If the companion mass exceeded $10\,M_{\rm
Jup}$, we drew the eccentricity distribution from the power-law $p_e
\propto e^\eta$ reported by \citet{moe_mind_2017} in their
Equation~17.  We emphasize that the long-period exoplanet and
long-period binary eccentricity distributions are quite different: the
exoplanet distribution is ``bottom-heavy'', with eccentricities
preferentially close to zero. The binary star distribution is
``top-heavy'', with eccentricities closer to one.  We chose a mass
cutoff of $10\,M_{\rm Jup}$ to separate the two regimes based on the
bound observed by \citet{schlaufman_evidence_2018} between giant
planets and brown dwarfs, though this value is also close to the
$13\,M_{\rm Jup}$ deuterium-burning limit \citep[{\it
e.g.},][]{burrows_nongray_1997}.

>>>>

A cursory analysis of the new HIRES observations led to our detection
of a linear trend in the residuals after fitting out the orbit of
WASP-4b.  This prompted us to acquire speckle images using Zorro at
Gemini-South \citep[see][and the instrument
web-pages\footnote{\url{www.gemini.edu/sciops/instruments/alopeke-zorro/}}]{scott_nessi_2018}.
Zorro is a dual-channel speckle interferometer employing narrow-band
filters centered at 562$\,$nm and 832$\,$nm.

We observed WASP-4 twice, on the night of September 11-12 with
relatively poor seeing (1.2$''$) and also on the night of September
28-29.  Over each observation, we acquired three sets of 1000
60$\,$msec exposures.  If a companion is present, the autocorrelation
functions of these speckle images reveal a characteristic
interference pattern. This pattern in then used to determine the
properties of the detected companion and to produce a reconstructed
image.  Using the reduced images and their
reconstructed speckle images,  contrast curves are produced yielding
5-sigma detection limits  (see \citealt{howell_speckle_2011}). 
No companions were detected,
and the second night, which had better seeing
(0.6$''$), also produced the more constraining result.  The 832$\,$nm
limits were the most useful to us given that faint companions
are redder than the host star.  We therefore opted to use the
832$\,$nm September 28-29 contrast limits for the remaining analysis.

@@

The system's acceleration towards our line-of-sight causes a decrease
in the apparent orbital period.  The expected period derivative is
\begin{align}
  \dot{P}_{\rm\,RV}
  = \frac{\dot{v}_{\rm r} P}{c}
  = -5.94 \pm 0.39~{\rm msec}\,{\rm yr}^{-1}.
\end{align}
In other words, the majority of the period decrease observed in
transits ($\dot{P}= - 8.64 \pm 0.89~{\rm msec}\,{\rm yr}^{-1}$) seems to
be explained by the acceleration of the host star.

@@

Given a linear radial velocity trend, we can place lower-limits on the
mass and semi-major axis of additional bodies in the system.  For a
quick estimate of the minimum mass required to explain the linear
trend in WASP-4, we turn to \citet{feng_california_2015}.  As they
discuss, the scenario that yields the minimum companion mass for a
system with a linear trend is a companion with $e\approx0.5$ and
$\omega=90^\circ$.  Substituting $P\approx 1.25\tau$ and $K \approx
0.5\tau \dot{\gamma}$ into the mass function \citep[{\it
e.g.},][]{wright_efficient_2009} yields
\begin{equation}
 M_{\rm min} \approx 5.99\,M_{\rm Jup}
  \left( \frac{\tau}{{\rm yr}} \right)^{4/3}
  \left| \frac{\dot{\gamma}}{{\rm m\,s^{-1}\,day^{-1}}} \right|
  \left( \frac{M_\star}{{M_\odot}} \right)^{2/3},
\end{equation}
where $\tau$ is the observing baseline.  For WASP-4, this gives
$M_{\rm min} = 4.9 M_{\rm Jup}$.  Higher masses are allowed for
companions that orbit further from the star: at fixed $\dot{\gamma}$,
$M_{\rm comp} \propto a^2$
\citep{torres_substellar_1999,liu_crossing_2002}.

@@

Due to the custom filters of the Zorro imager, and corresponding lack
of synthetic photometry, we further assumed that all sources had
blackbody spectra. While this is a simplification, we do not
readily have access to the planetary and stellar atmosphere models
needed for the consistent calculation with the \texttt{COND03} and
\texttt{MESA} models.  We therefore adopted the effective temperatures
and bolometric luminosities from the \citet{baraffe_evolutionary_2003}
and MIST isochrones.  Using these theoretical quantities and the
empirically-measured Zorro bandpasses, we calculated absolute
magnitudes in the 562 and 832 nm Zorro bands for stellar and planetary
mass companions.  Applying the same calculation to WASP-4 itself using
the effective temperature and bolometric luminosity from
\citetalias{bouma_wasp4b_2019}, we derived the transformation from
contrast ratio to companion mass.  The resulting limits are shown in
Figure~\ref{fig:mass_sma}.

@@

We assigned each companion a mass and semimajor axis from log-uniform
distributions within the grid cell. We drew the inclination from a
uniform distribution in $\cos i$.  For companion masses less than
$10\,M_{\rm Jup}$, we drew the eccentricity from
\citet{kipping_beta_2013}'s long-period exoplanet Beta distribution
($a=1.12$, $b=3.09$).  If the companion mass exceeded $10\,M_{\rm
Jup}$, we drew the eccentricity from the power-law $p_e \propto
e^\eta$ reported by \citet{moe_mind_2017} in their Equation~17 ($\eta
\approx 0.5$ for most orbital periods).  The long-period exoplanet and
long-period binary eccentricity distributions are quite different: the
exoplanet distribution is ``bottom-heavy'', with eccentricities
preferentially close to zero.  The binary star distribution is
``top-heavy'', with eccentricities closer to one.  We chose a mass
cutoff of $10\,M_{\rm Jup}$ to separate the two regimes based on the
bound observed by \citet{schlaufman_evidence_2018} between giant
planets and brown dwarfs, though this value is also close to the
$13\,M_{\rm Jup}$ deuterium-burning limit \citep[{\it
e.g.},][]{burrows_nongray_1997}.
